---
title: "CMPU4091 Visualising Data"
subtitle: "Correlation"
author: "Matthew Bradon C21380786"
format:
  html:
    code-fold: true
    code-tools: true
    css: "./styles.css"
editor: visual
---

# Introduction

<p>

The survey.dat dataset contains survey responses about their mental well-being. It includes information like age, gender, smoking habits, life satisfaction, and optimism.

This document outlines the relationship between positive and negative affect, mastery and smoking habits in relation to optimism.

</p>

```{r setup, include="FALSE"}

#Setting include to FALSE on this chunk so that it doesn't appear in the output
# Load necessary libraries
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("tidyverse")) install.packages("tidyverse", dependencies = TRUE)
if (!require("cowplot")) install.packages("cowplot", dependencies = TRUE)
if (!require("plotly")) install.packages("plotly", dependencies = TRUE)
if (!require("ggcorrplot")) install.packages("ggcorrplot", dependencies = TRUE)
if (!require("pastecs")) install.packages("pastecs", dependencies = TRUE)
if (!require("PerformanceAnalytics")) install.packages("PerformanceAnalytics", dependencies = TRUE)
if (!require("patchwork")) install.packages("patchwork", dependencies = TRUE)
if (!require("Hmisc")) install.packages("Hmisc", dependencies = TRUE)
if (!require("corrplot")) install.packages("corrplot", dependencies = TRUE)
if(!require("mgcv")) install.packages("mgcv", dependencies = TRUE)


library(tidyverse) # Tools for wrangling data
library(cowplot) # For organising plots into a grid
library(patchwork) # For matrix layout
library(plotly) # For Animating plots
library(ggcorrplot) # For Correlation Heatmaps 
library(corrplot) # For correlation matrices
library(PerformanceAnalytics)# For standardised skewness and kurtosis
library(Hmisc) # For generating correlation


datapath <- file.path('survey.dat')

#Load it into the dataframe called survey (.dat file so use read.Table)
survey <- read.table(datapath)

#Setting the column names to be that used in the dataset but in lowercase to make life a bit easier
colnames(survey) <- tolower(colnames(survey))

# Omit Null values
survey <- na.omit(survey)
```

# Histograms with Density Plots

```{r}
# Optimism Histogram
toptim_hist <- ggplot(survey, aes(x = toptim)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, alpha = 0.4, position = "identity", fill = "red") +
  geom_density(alpha = 0.7, adjust = 1.2) +
  labs(title = "Optimism", x = "Optimism", y = "Density") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 10, 5, 5)
  )

# Positive Affect Histogram
tposaff_hist <- ggplot(survey, aes(x = tposaff)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, alpha = 0.4, position = "identity", fill = "blue") +
  geom_density(alpha = 0.7, adjust = 1.2) +
  labs(title = "Positive Affect", x = "Positive Affect", y = "Density") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 10, 5, 5)
  )

# Negative Affect Histogram
tnegaff_hist <- ggplot(survey, aes(x = tnegaff)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, alpha = 0.4, position = "identity", fill = "darkgreen") +
  geom_density(alpha = 0.7, adjust = 1.2) +
  labs(title = "Negative Affect", x = "Negative Affect", y = "Density") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 10, 5, 5)
  )

# Mastery Histogram
tmast_hist <- ggplot(survey, aes(x = tmast)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, alpha = 0.4, position = "identity", fill = "purple") +
  geom_density(alpha = 0.7, adjust = 1.2) +
  labs(title = "Mastery", x = "Mastery", y = "Density") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 10, 5, 5)
  )

# Arrange the plots using cowplot
histogram_grid <- plot_grid(toptim_hist, tposaff_hist, tnegaff_hist, tmast_hist, nrow = 2, ncol = 2)

# Display the grid
print(histogram_grid)
```

<p>Optimism, Postive affect and Mastery are left skewed and Negative Affect is right skewed.</p>

# Smoking Status Bar Chart

```{r}
smoke_bar <- ggplot(survey, aes(x = smoke, fill = as.factor(smoke))) +  # Color by category
  geom_bar() +
  labs(title = "Bar Chart of Smoking Status", x = "Smoking Status", y = "Count", fill = "Smoking Status") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 10, 5, 5)
  )

print(smoke_bar)
```

<p>There is a major class imbalance of non smoker to smokers is this a possible source of the skewness we observe above.</p>

# Box plot Comparing optimism between smokers and non smokers

```{r}
# Box plot Comparing optimism between smokers and non smokers
smoke_box <- ggplot(survey, aes(x = smoke, y = toptim, fill = smoke)) +
  geom_boxplot() +
  labs(title = "Boxplot of Optimism by Smoking Status", x = "Smoking Status", y = "Optimism", fill = "Smoking Status") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(5, 10, 5, 5)
  )

print(smoke_box)
```

<p>

The 3rd quartile of people who dont smoke is higher but the median of smoker is higher than non smokers. The 1st quartile of smokers is lower than that of non smokers

The minimum and maximum of both classes are around the same

In non smokers there are two outliers who are really low on optimism.

We do have significantly less examples of smokers than non smokers if our dataset was larger there might have been a bigger difference in the box plots.

</p>

# Exploring Relationships

## Correlation coefficients and p-values

<p>The correlation between Total optimism and positive affect, negative affect and mastery</p>

```{r}
for (var in c("tposaff", "tnegaff", "tmast")) {
  var_data <- survey %>% select(toptim, !!sym(var)) %>% 
    rename(
      Optimism = toptim,
      Variable = !!sym(var)
    )
  
  if (nrow(var_data) > 1) {
    cor_result <- cor.test(var_data$Optimism, var_data$Variable)
    cat(paste("\nVariable:", var, "\n"))
    print(data.frame(
      Variable = var,
      Correlation = cor_result$estimate,
      P_value = cor_result$p.value
    ))
  }
}
```

<p>

Each of the correlations are moderate i.e 0.3 to 0.6

Mastery has the highest correlation with Optimism at 0.569.

All of the correlations for the psychological factors all have a P value \< 0.05 making it seem to be statically significant enough to rule out random chance.

</p>

## Scatterplot with linear regression

```{r}
p1 <- ggplot(survey, aes(x = tposaff, y = toptim)) +
    geom_point(alpha = 0.5, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = "Optimism vs. Positive Affect", x = "Positive Affect", y = "Optimism") +
    theme_minimal()

p2 <- ggplot(survey, aes(x = tnegaff, y = toptim)) +
    geom_point(alpha = 0.5, color = "green") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = "Optimism vs. Negative Affect", x = "Negative Affect", y = "Optimism") +
    theme_minimal()

p3 <- ggplot(survey, aes(x = tmast, y = toptim)) +
    geom_point(alpha = 0.5, color = "purple") +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(title = "Optimism vs. Mastery", x = "Mastery", y = "Optimism") +
    theme_minimal()

# Combine plots using patchwork
scatter_plots <-  (p1 / p2 / p3) + 
    patchwork::plot_annotation(title = "Scatterplots of Optimism and Psychological Factors", theme = theme(plot.title = element_text(hjust = 0.5)))


print(scatter_plots)

```

<p>

Positive affect has a a moderate positive correlation with Optimism meaning that happier people tend to be more optimistic

Mastery has a moderate positive correlation with Optimism meaning those who feel in control tend to be more optimistic

Negative Affect has a moderate negative correlation meaning that more negative emotions tend to lower your optimism

</p>

## Correlation Heatmap

```{r}
# Select only relevant columns
selected_data <- survey %>% select(toptim, tposaff, tnegaff, tmast)

# Rename for better labels
colnames(selected_data) <- c("Optimism", "Positive Affect", "Negative Affect", "Mastery")

# Compute correlation matrix
corr_matrix <- cor(selected_data, use = "complete.obs")

# Generate heatmap
p <- ggcorrplot(corr_matrix, 
              type = "lower", 
              lab = TRUE, 
              lab_size = 3, 
              colors = c("#6D9EC1", "white", "#E46726"), 
              title = "Correlation Heatmap of Psychological Factors", 
              ggtheme = theme_minimal()) +
theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))


print(p)
```

<p>

Blue squares are negative correlation and Red squares are positive correlation

I.e Positive Affect and Negative Affect have moderate negative correlation as happier people tend to have less negative thoughts.

People who are more in control tend to be happier and optimistic

</p>