---
title: "L11-Visualizing Text"
format: 
  html:
    self-contained: true
    code-fold: true
    toc: true
    toc_float: true
    toc-location: left
editor: source
---


This document demonstrates text visualization approaches.

```{r setup}
# Set the CRAN repository
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Function to quietly install and load packages
quiet_library <- function(pkg) {
  suppressMessages(suppressWarnings({
    if (!requireNamespace(pkg, quietly = TRUE)) {
      install.packages(pkg, quiet = TRUE) # Install without messages
    }
    library(pkg, character.only = TRUE) # Load silently
  }))
}

quiet_library("tidyverse")        # Core packages for data wrangling, visualization (includes dplyr, ggplot2, etc.)

quiet_library("tidytext")         # Text mining using tidy principles; unnest_tokens(), sentiment analysis, TF-IDF

quiet_library("tm")               # Traditional text mining tools for corpus cleaning (e.g., removePunctuation(), stopwords)

quiet_library("text2vec")         # High-performance text vectorization (word embeddings, TF-IDF, GloVe, etc.)

quiet_library("tokenizers")       # Fast and flexible tokenization (words, sentences, ngrams, etc.)

quiet_library("Rtsne")            # t-SNE algorithm for dimensionality reduction (e.g., 2D projections of word vectors)

quiet_library("igraph")           # Graph and network analysis (used for word co-occurrence networks, semantic graphs)

quiet_library("ggraph")           # Visualization of graphs and networks (works with igraph; great for word trees and clusters)

quiet_library("ggplot2")          # Grammar of graphics plotting system (used to visualize frequency, sentiment, etc.)

quiet_library("widyr")            # Pairwise correlations and distances between words (e.g., word co-occurrence)

quiet_library("SnowballC")        # Stemming (reducing words to root form, e.g., "running" → "run")

quiet_library("wordcloud")        # Generates word cloud visualizations based on term frequency

quiet_library("RColorBrewer")     # Provides color palettes for use in plots (especially useful in wordclouds and ggplot2)

quiet_library("pdftools")         # Reads text content from PDF files


quiet_library("collapsibleTree")  # Creates interactive, collapsible D3-style trees (e.g., for word trees, suffix trees)

# Text data to use
text_data <- c(
  "Data Science is fun and useful.",
  "Machine learning is challenging."
)
```


# Ngrams
## Bigrams
```{r ngram}
# Extract bigrams (n = 2)
bigrams <- tokenizers::tokenize_ngrams(text_data, n = 2)
print(bigrams)

# Using tidytext
df <- tibble::tibble(text = text_data)

bigrams_df <- df %>%
  tidytext::unnest_tokens(bigram, text, token = "ngrams", n = 2)

print(bigrams_df)

```
## Trigrams
```{r tgram, message=FALSE, warning=FALSE}
# Extract trigrams (n = 3)
trigrams <- tokenizers::tokenize_ngrams(text_data, n = 3)
print(trigrams)

trigrams_df <- df %>%
  tidytext::unnest_tokens(trigram, text, token = "ngrams", n = 3)

print(trigrams_df)

```

# Bag of Words
```{r bow, message=FALSE, warning=FALSE}
text_data <- c(
  "Data Science is fun and useful.",
  "Machine learning is challenging."
)

# Create a corpus object using the 'tm' package
corpus <- tm::Corpus(tm::VectorSource(text_data))

# Preprocess the corpus using 'tm' functions
corpus <- tm::tm_map(corpus, tm::content_transformer(tolower))
corpus <- tm::tm_map(corpus, tm::removePunctuation)
corpus <- tm::tm_map(corpus, tm::removeWords, tm::stopwords("en"))
corpus <- tm::tm_map(corpus, tm::stripWhitespace)

# Create a Document-Term Matrix (DTM)
dtm <- tm::DocumentTermMatrix(corpus)

# Convert DTM to a matrix
bow_matrix <- as.matrix(dtm)

# Convert to a data frame and print
bow_df <- as.data.frame(bow_matrix)
print(bow_df)

```

# TD-IDF
```{r tdidf, message=FALSE, warning=FALSE}
# Create a corpus
corpus <- tm::Corpus(tm::VectorSource(text_data))

# Preprocess: lowercase, remove punctuation and stopwords, strip whitespace
corpus <- tm::tm_map(corpus, tm::content_transformer(tolower))
corpus <- tm::tm_map(corpus, tm::removePunctuation)
corpus <- tm::tm_map(corpus, tm::removeWords, tm::stopwords("en"))
corpus <- tm::tm_map(corpus, tm::stripWhitespace)

# Create Document-Term Matrix
dtm <- tm::DocumentTermMatrix(corpus)

# Convert to TF-IDF
dtm_tfidf <- tm::weightTfIdf(dtm)

# Convert to a matrix and print
tfidf_matrix <- as.matrix(dtm_tfidf)


print(tfidf_matrix)
```

# Stop Word Removal
## Stop Word List
```{r sw1, message=FALSE, warning=FALSE}
# Using 'tm' for preprocessing
corpus <- tm::Corpus(tm::VectorSource(text_data))

# Convert to lowercase, remove punctuation, and stop words
corpus <- tm::tm_map(corpus, tm::content_transformer(tolower))
corpus <- tm::tm_map(corpus, tm::removePunctuation)
corpus <- tm::tm_map(corpus, tm::removeWords, tm::stopwords("en"))  # Common English Stop Words
corpus <- tm::tm_map(corpus, tm::stripWhitespace)

# Inspect cleaned text
tm::inspect(corpus)

# tidytext approach
df <- tibble::tibble(text = text_data)

# Tokenize and Remove Stop Words using tidytext
cleaned_df <- df %>%
  tidytext::unnest_tokens(word, text) %>%
  dplyr::anti_join(tidytext::stop_words, by = "word")

print(cleaned_df)
```

## Using Tokenizers
```{r sw2, message=FALSE, warning=FALSE}
# Tokenize using the 'tokenizers' package
tokens <- tokenizers::tokenize_words(text_data)

# Remove stop words manually using the 'tm' package
filtered_tokens <- lapply(tokens, function(words) {
  words[!words %in% tm::stopwords("en")]
})

print(filtered_tokens)

# Repeat (if needed)
tokens <- tokenizers::tokenize_words(text_data)

filtered_tokens <- lapply(tokens, function(words) {
  words[!words %in% tm::stopwords("en")]
})


print(filtered_tokens)
```


# Word Embedding - Semantic Similarity
## Table
```{r word2vec, message=FALSE, warning=FALSE}
# Tokenize - break text_data into individual words
tokens_iterator <- text2vec::itoken(text_data, tokenizer = text2vec::word_tokenizer)

# Create the vocabulary and vectorizer - builds a list of unique terms
vocab <- text2vec::create_vocabulary(tokens_iterator) %>%
  text2vec::prune_vocabulary(term_count_min = 1)
vectorizer <- text2vec::vocab_vectorizer(vocab)

# Create the term co-occurrence matrix (TCM)
# Shows how frequently terms appear close to each other
tcm <- text2vec::create_tcm(tokens_iterator, vectorizer, skip_grams_window = 3)

# Create and train the GloVe model
# Learns word embeddings: dense vector representations of words that capture semantic meaning
glove_model <- text2vec::GloVe$new(rank = 50, x_max = 10)  # Create the model
word_vectors <- glove_model$fit_transform(tcm, n_iter = 10)  # Fit and transform

print(word_vectors)
```
*Table: Each row represents a word, and each column corresponds to a dimension in the 50-dimensional semantic space.

## Visualization - Show how words cluster based on meaning
```{r cluster, , message=FALSE, warning=FALSE}
# Run t-SNE to reduce 50-dimensional word vectors to 2D for visualization.
# Perplexity is set to 2 to ensure it's suitable for the small number of words (must be < n/3) where n is the number of rows in word_vectors.
# Run t-SNE to reduce word vectors to 2D
tsne_model <- Rtsne::Rtsne(word_vectors, dims = 2, perplexity = 2)

# Plot the result
graphics::plot(tsne_model$Y, type = "n")
graphics::text(tsne_model$Y, labels = rownames(word_vectors), cex = 0.7)

```
# Semantic Network
```{r semnet, message=FALSE, warning=FALSE}
# Turn into a tibble
text_df <- tibble::tibble(doc_id = 1:length(text_data), text = text_data)

# Tokenize and clean
tokens <- text_df %>%
  tidytext::unnest_tokens(word, text) %>%
  dplyr::filter(!word %in% tidytext::stop_words$word)  # remove stopwords

# Create word pairs (bigrams within each doc)
word_pairs <- tokens %>%
  widyr::pairwise_count(word, doc_id, sort = TRUE, upper = FALSE)

# Build graph
graph <- igraph::graph_from_data_frame(word_pairs)

# Visualize
ggraph::ggraph(graph, layout = "fr") +  # fr = force-directed layout
  ggraph::geom_edge_link(ggplot2::aes(edge_alpha = n), show.legend = FALSE) +
  ggraph::geom_node_point(color = "steelblue", size = 5) +
  ggraph::geom_node_text(ggplot2::aes(label = name), repel = TRUE) +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Semantic Network of Word Co-occurrences")

```

# Stemming
```{r stem, message=FALSE, warning=FALSE}
# Sample text data
text_data <- c(
  "Natural language processing is fascinating.",
  "Text mining involves many preprocessing steps.",
  "Tokenization and stemming help analyze text."
)

# Create a data frame
text_df <- tibble::tibble(line = 1:3, text = text_data)

# Tokenize
tokens <- text_df %>%
  tidytext::unnest_tokens(word, text)

# Remove stopwords
data(tidytext::stop_words)  # built-in list of stopwords
tokens_clean <- tokens %>%
  dplyr::filter(!word %in% tidytext::stop_words$word)

# Apply stemming (SnowballC)
tokens_clean <- tokens_clean %>%
  dplyr::mutate(stem = SnowballC::wordStem(word))

# Count word frequency
word_counts <- tokens_clean %>%
  dplyr::count(stem, sort = TRUE)

# View top words
print(word_counts)

# Visualize top stems
word_counts %>%
  dplyr::top_n(10) %>%
  ggplot2::ggplot(ggplot2::aes(x = stats::reorder(stem, n), y = n)) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::coord_flip() +
  ggplot2::labs(title = "Most Common Word Stems",
                x = "Stemmed Word", y = "Frequency") +
  ggplot2::theme_minimal()

```

# Word Cloud
```{r wcloud, message=FALSE, warning=FALSE}
# Step 1: Read text from the PDF
pdf_text_data <- pdftools::pdf_text("C://datasets//VisualizingDataModuleSpec.pdf")

# Step 2: Combine pages into one text blob
full_text <- paste(pdf_text_data, collapse = " ")

# Step 3: Create a text corpus
corpus <- tm::Corpus(tm::VectorSource(full_text))

# Step 4: Preprocess the text
corpus <- tm::tm_map(corpus, tm::content_transformer(tolower))         # Lowercase
corpus <- tm::tm_map(corpus, tm::removePunctuation)                    # Remove punctuation
corpus <- tm::tm_map(corpus, tm::removeNumbers)                        # Remove numbers
corpus <- tm::tm_map(corpus, tm::removeWords, tm::stopwords("en"))    # Remove English stopwords
corpus <- tm::tm_map(corpus, tm::stripWhitespace)                      # Remove extra whitespace

# Step 5: Create a term-document matrix
tdm <- tm::TermDocumentMatrix(corpus)
matrix <- as.matrix(tdm)

# Step 6: Calculate word frequencies
word_freqs <- sort(rowSums(matrix), decreasing = TRUE)
word_df <- data.frame(word = names(word_freqs), freq = word_freqs)

# Step 7: Generate the word cloud
set.seed(123)
wordcloud::wordcloud(
  words = word_df$word,
  freq = word_df$freq,
  min.freq = 2,
  max.words = 100,
  random.order = FALSE,
  colors = RColorBrewer::brewer.pal(8, "Dark2")
)


```
# Word Tree
## Word Tree (approximation)
```{r wtree, message=FALSE, warning=FALSE}
# Convert to tibble and tokenize
text_df <- tibble::tibble(text = full_text)

tokens <- text_df %>%
  tidytext::unnest_tokens(word, text, token = "words")

# Create word pairs (bigrams)
bigrams <- text_df %>%
  tidytext::unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  tidyr::separate(bigram, into = c("word1", "word2"), sep = " ") %>%
  dplyr::filter(!is.na(word2))

# Optional: filter to root word (e.g., build branches from "data")
root_word <- "data"
branches <- bigrams %>%
  dplyr::filter(word1 == root_word)

# Create graph object
graph <- igraph::graph_from_data_frame(branches)

# Visualize as a tree (word tree approximation)
ggraph::ggraph(graph, layout = "tree") +
  ggraph::geom_edge_link() +
  ggraph::geom_node_point(color = "steelblue", size = 3) +
  ggraph::geom_node_text(ggplot2::aes(label = name), vjust = -0.5, hjust = 0.5) +
  ggplot2::theme_void() +
  ggplot2::labs(title = paste("Word Tree Rooted at:", root_word))

```
## Collapsible D3 Style Word Tree
```{r D3, message=FALSE, warning=FALSE}

# Sample data
sample <- c(
  "cats are better than dogs",
  "cats eat kibble",
  "cats are better than hamsters",
  "cats are awesome",
  "cats are people too",
  "cats eat mice",
  "cats meowing",
  "cats in the cradle",
  "cats eat mice",
  "cats in the cradle lyrics",
  "cats eat kibble",
  "cats for adoption",
  "cats are family",
  "cats eat mice",
  "cats are better than kittens",
  "cats are evil",
  "cats are weird",
  "cats eat mice"
)
# Tokenize into individual words and build a data frame where each column is a level
tokens_df <- tibble::tibble(sentence = sample) %>%
  dplyr::mutate(id = dplyr::row_number()) %>%
  tidyr::separate_rows(sentence, sep = " ") %>%
  dplyr::filter(!sentence %in% tidytext::stop_words$word) %>% 
  dplyr::group_by(id) %>%
  dplyr::mutate(level = paste0("level_", dplyr::row_number())) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = level, values_from = sentence)

# View a few rows
head(tokens_df)

# Visualize as collapsible D3 tree
collapsibleTree::collapsibleTree(
  tokens_df,
  hierarchy = names(tokens_df)[-1],  # exclude the ID column
  root = "cat hierarchy",
  width = 800,
  height = 600,
  fontSize = 16,
  zoomable = TRUE
)


```


# Differences in Documents

## Word Cloud per document
```{r wcdiff, message=FALSE, warning=FALSE}
# Your custom stopwords
custom_stopwords <- c("marks", "cmpu", "4091", "semester", "exam", "supplemental", "dt228", "dt211c", "dt282", "tu856", "tu857", "tu858" ,"cmpu4091", "question", "paper", "examination", "figure", "e.g", "questions", "summer", "honours", "degree", "computer", "science", "international", "data", "visualizing", "visualization", "explain", "describe", "example", "instructions", "duration", "can", "internal", "external", "examiners", "date", "percentage", "bsc", "dr", "ms", "mr", "may")

# Combine with built-in stopwords
all_stopwords <- c(stop_words$word, custom_stopwords)


# Read PDF text (update the file paths if needed)
pdf1 <- pdftools::pdf_text("C://datasets//Exam2023.pdf")
pdf2 <- pdftools::pdf_text("C://datasets//Exam2024.pdf")
pdf3 <- pdftools::pdf_text("C://datasets//ExamSupp2024.pdf")

# Combine all pages into one string per document
text_data <- list(
  "Exam 2023" = paste(pdf1, collapse = " "),
  "Exam 2024" = paste(pdf2, collapse = " "),
  "Supplemental 2024" = paste(pdf3, collapse = " ")
)

# Convert to a corpus
corpus <- tm::Corpus(tm::VectorSource(text_data))

# Clean and preprocess
corpus <- tm::tm_map(corpus, tm::content_transformer(tolower))                            # Lowercase
corpus <- tm::tm_map(corpus, tm::removePunctuation)                                       # Remove punctuation
corpus <- tm::tm_map(corpus, tm::removeNumbers)                                           # Remove numbers
corpus <- tm::tm_map(corpus, tm::removeWords, c(tm::stopwords("en"), custom_stopwords))  # Remove stopwords
corpus <- tm::tm_map(corpus, tm::stripWhitespace)                                         # Remove extra spaces

# Set layout to show all three word clouds side by side
graphics::par(mfrow = c(1, 3))  # 1 row, 3 columns

# Loop through each document in the corpus and generate a word cloud
for (i in 1:length(corpus)) {
  # Create term-document matrix
  tdm <- tm::TermDocumentMatrix(corpus[i])
  m <- as.matrix(tdm)
  word_freqs <- sort(rowSums(m), decreasing = TRUE)
  
  # Generate word cloud
  wordcloud::wordcloud(
    words = names(word_freqs),
    freq = word_freqs,
    min.freq = 2,
    max.words = 100,
    random.order = FALSE,
    colors = RColorBrewer::brewer.pal(8, "Dark2")
  )
}


```


## Bar Chart of Word Frequency

```{r bchart, message=FALSE, warning=FALSE}
# Combine each PDF into a single string
doc_data <- tibble::tibble(
  doc_id = c("Exam 2023", "Exam 2024", "Supplemental 2024"),
  text = c(
    paste(pdf1, collapse = " "),
    paste(pdf2, collapse = " "),
    paste(pdf3, collapse = " ")
  )
)

# Tokenize and clean data
data_tokens <- doc_data %>%
  tidytext::unnest_tokens(word, text) %>%
  dplyr::filter(!word %in% all_stopwords) %>%
  dplyr::filter(stringr::str_detect(word, "[a-z]"))  # optional: only alphabetic

# Count words per document
word_counts <- data_tokens %>%
  dplyr::count(doc_id, word, sort = TRUE)

# Get top 10 words per document
top_words <- word_counts %>%
  dplyr::group_by(doc_id) %>%
  dplyr::slice_max(order_by = n, n = 10) %>%
  dplyr::ungroup()

# Plot the top words
ggplot2::ggplot(top_words, ggplot2::aes(x = tidytext::reorder_within(word, n, doc_id), y = n, fill = doc_id)) +
  ggplot2::geom_col(show.legend = FALSE) +
  ggplot2::facet_wrap(~ doc_id, scales = "free_y") +
  tidytext::scale_x_reordered() +
  ggplot2::coord_flip() +
  ggplot2::labs(
    title = "Top 10 Most Frequent Words per Exam Document",
    x = "Word",
    y = "Frequency"
  ) +
  ggplot2::theme_minimal()


```

